---
title: "Usage guide"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide shows zutoml's main tasks, one section each. For a tour built
around real configuration files, see `vignette("zutoml")`.

```{r setup}
library(zutoml)
```

## Read a document

`toml_parse()` reads TOML text; `toml_read()` reads a file, a URL or a
connection. Both return a named list, with keys in the order the document
gives them.

```{r}
doc <- '
title = "Example"

[server]
host = "localhost"
port = 8080
tags = ["web", "api"]

[[users]]
name = "ana"
admin = true

[[users]]
name = "bo"
'
x <- toml_parse(doc)
str(x)
```

```{r}
path <- tempfile(fileext = ".toml")
writeLines(doc, path)
toml_read(path)$server$port
```

## Read only part of a document

`select` returns the value at a path, written as the TOML key syntax:
dots between keys, quotes around keys that need them, and `[i]` for array
elements, counting from 1.

```{r}
toml_parse(doc, select = "server")
toml_parse(doc, select = "users[2].name")
toml_read(path, select = c("server", "tags"))
```

The whole document is still checked, since TOML lets a table be added to
anywhere in it, but only the selected part becomes R values. A path the
document does not have is an error, never `NULL`:

```{r}
try(toml_parse(doc, select = "server.timeout"))
```

## Types

Every TOML type maps to one R type, chosen by the document's syntax, never
by what a string looks like:

| TOML | R |
|---|---|
| string | `character` |
| integer | `integer`; `double` past R's integer range; `toml_bigint` past 2^53 |
| float | `double` (`inf`, `nan` as `Inf`, `NaN`) |
| boolean | `logical` |
| offset date-time | `POSIXct` in UTC |
| local date-time | `POSIXct` in the session's time zone |
| local date | `Date` |
| local time | `character`, or `difftime` |
| table | named `list` |
| array | a vector when its elements share a type, else a `list` |

```{r}
str(toml_parse('
when  = 1979-05-27T07:32:00-08:00
day   = 1979-05-27
alarm = 07:30:00
text  = "1979-05-27"
big   = 9223372036854775807
'))
```

`datetimes = "keep"` returns date-times as RFC 3339 text, and
`local_time = "difftime"` returns times of day as seconds since midnight.
`big_integers` decides what happens past 2^53: a `toml_bigint` (exact
text, the default), the nearest double, or an error.

## Arrays

An array whose elements share a type becomes a vector; integers and floats
together become a `double` vector. Booleans never join the numbers.
Anything else is a list.

```{r}
str(toml_parse("a = [1, 2]\nb = [1, 2.5]\nc = [true, 1]\nd = []\ne = [3]"))
```

A one-element array is marked with `I()` so that writing it back gives an
array again, as `e` shows. `simplify = "none"` makes every array a list.

## Data frames

`data_frame = TRUE` turns an array of tables into a data frame: one row per
table, columns in first-seen key order, `NA` where a table lacks a key.

```{r}
toml_parse(doc, data_frame = TRUE)$users
```

## Write a document

`toml_emit()` writes a named list as TOML text; `toml_write()` writes it to
a file or connection.

```{r}
cfg <- list(
  title = "Example",
  port = 8080,
  ratio = 0.75,
  started = as.Date("2026-10-08"),
  server = list(host = "localhost", tags = c("web", "api")),
  users = list(list(name = "ana"), list(name = "bo"))
)
cat(toml_emit(cfg))
```

- A table's own values come first, then its sub-tables, as TOML requires.
- `port = 8080` is a double in R but is written as the integer a TOML
  reader expects; non-whole doubles are written in the shortest digits that
  read back exactly.
- A list of named lists, or a data frame, becomes an array of tables.
- The output is the same on every platform, and reads back:

```{r}
identical(toml_parse(toml_emit(cfg))$users, cfg$users)
```

`NA` and `NULL` have no TOML form, so they are refused unless
`na = "omit"`. Arrays past `width` characters are written one element per
line, and `inline = n` writes small flat tables inline.

## Infrastructure configuration

Many infrastructure tools are configured in TOML: Traefik, containerd,
GitLab Runner, Telegraf, Netlify, Cloudflare's `wrangler.toml`, Fly.io's
`fly.toml`. Reading one is the same as reading any document, here Traefik's
sample static configuration, pinned to a release:

```{r}
traefik <- "https://raw.githubusercontent.com/traefik/traefik/v3.7.14/traefik.sample.toml"
static <- toml_read(traefik)
names(static)
static$entryPoints$websecure
```

Writing is where TOML from R pays off: a configuration generated from data.
This is a Traefik dynamic configuration that routes a host to a set of
backends:

```{r}
backends <- c("http://10.0.0.1:8080", "http://10.0.0.2:8080")
dynamic <- list(http = list(
  routers = list(app = list(
    rule = "Host(`app.example.com`)",
    service = "app",
    entryPoints = I("websecure"),
    tls = list(certResolver = "letsencrypt")
  )),
  services = list(app = list(
    loadBalancer = list(servers = lapply(backends, function(u) list(url = u)))
  ))
))
cat(toml_emit(dynamic))
```

`I("websecure")` keeps a one-element array an array, which Traefik expects;
the list of servers becomes an array of tables. Write it where Traefik
watches for files with `toml_write(dynamic, "dynamic.toml")`.

## Choose how one value is written

Markers change how a value is written, never what it reads back as:

```{r}
cat(toml_emit(list(
  dependencies = list(
    serde = toml_inline(list(version = "1.0", features = "derive")),
    rand = "0.8"
  ),
  pattern = toml_literal("^\\d+\\.\\d+$"),
  motd = toml_multiline("Welcome!")
)))
```

`toml_inline()` writes a table inline, `toml_literal()` writes literal
strings that need no escapes, and `toml_multiline()` writes multi-line
strings.

## Point at a value in the document

`positions = TRUE` records where every key and array element is defined,
so a check on a configuration can say where the problem is:

```{r}
cfg <- toml_parse("[server]\nhost = 'localhost'\nport = 99999\n", positions = TRUE)
pos <- attr(cfg, "toml_positions")
pos
at <- pos[pos$path == "server.port", ]
sprintf("server.port (line %d, column %d) must be at most 65535", at$line, at$column)
```

The `path` column uses the same syntax as `select`.

## Check a document

`toml_validate()` runs the whole check without building any R value. It
returns `TRUE` or `FALSE`, or with `error = TRUE` raises the reason:

```{r}
toml_validate("a = 1\nb = 2")
toml_validate("a = 1\na = 2")
try(toml_validate("a = 1\na = 2", error = TRUE))
```

## Handle errors

Every error has a class, so it can be caught by kind. Parse errors carry the
`line`, `column`, byte `offset` and `kind` of the fault:

```{r}
err <- tryCatch(toml_parse("[t]\nx = 1\n[t]\n"), zutoml_parse_error = function(e) e)
c(err$kind, err$line, err$column)
```

| Class | When |
|---|---|
| `zutoml_parse_error` | the document is not TOML |
| `zutoml_limit_error` | a limit was reached (`zutoml_size_limit`, `zutoml_depth_limit`, `zutoml_item_limit`, `zutoml_string_limit`) |
| `zutoml_unrepresentable` | valid TOML that R cannot hold, such as a string with U+0000 |
| `zutoml_missing_key` | `select` names a path the document does not have |
| `zutoml_unsupported_type`, `zutoml_na_error` | an R value with no TOML form, or `NA`/`NULL` |
| `zutoml_invalid_argument`, `zutoml_io_error` | a bad argument; a file that cannot be read or written |

All of them inherit from `zutoml_error`. See `?"zutoml-conditions"`.

## Untrusted input

Four limits bound what a document can make zutoml do, and are checked
before any R value is built:

```{r}
try(toml_parse(paste0("a = ", strrep("[", 200)), max_depth = 64))
```

| Argument | Default | Bounds |
|---|---|---|
| `max_size` | 64 MiB | the document's bytes; `toml_read()` never reads past it |
| `max_depth` | 128 | nested tables and arrays (at most 1023) |
| `max_items` | 1e6 | keys plus array elements |
| `max_string` | 2^31 - 1 | one string's or key's bytes |

## TOML versions

zutoml reads TOML 1.1.0 by default, which adds `\e` and `\xHH` escapes,
times without seconds, and newlines and trailing commas in inline tables.
`version = "1.0.0"` reads strictly. The emitter always writes text that
TOML 1.0.0 readers accept.

```{r}
toml_parse("t = 07:30")$t
try(toml_parse("t = 07:30", version = "1.0.0"))
```
